Semantic information inside sentence embeddings

### 2023-05-13 paper/paper.tex
Rewrite probing tasks with new terminology.

Transformation \Trans{future} ....

@@

% TODO: doc2vec training - reread
For training the Doc2vec model, we have used the implementation of Doc2vec
algorithm available in the Python package gensim, using a window of size 4 and
considering every word that occured at all (no pruning of rare words) and the
vector size of 512. We have also tried training Doc2Vec models with different
embedding sizes (2, 3, 10, 100, 256), partly so that we could see, ehat impact
the size change has on the semantic label predictability, partly (especially if
we detected a positive correlation between the embedding size and the semantic
label predictability) in order to try and get hint as to how prominent the
specific semantic features are (i.e. which is the smallest vector size at which
given semantic label is predictable from the corresponding embeddings).

@@

The motivation for these tasks is that if the embedding contains semantic
information (in our case the type of transformation used), it would be
beneficial for the classifier to find and use this information. This would
boost its scores and make it standout.

We test two settings: \emph{no-context} and \emph{contextualized}. In
no-context setting the classifier receives only the sentence embedding itself
and test which (if any) transformations can be reliably determined without the
knowledge of the seed sentence. In the contextualized setting we pass both the
sentence embedding and the embedding of the corresponding seed sentence. This
enables the classifier to examine the relationships between the transformed and
seed sentences. The results are presented in Section~\ref{sec:probing_results},
the complete method is described in detail in
Appendix~\ref{appendix:probing_method}.

@@

\subsubsection{Contextualized setting}

We tried two methods of contextualization: \emph{concatenating} the two
embeddings and \emph{subtracting} the seed embedding from the sentence
embedding. Since the method using difference performed better, we omit the
concatenation results.

@@

  \caption{Grid search F1 macro averaged scores of contextualized embeddings.
  The reported scores of \Embed{sbert\_supervised} are the mean scores of all
  SBERT supervised embeddings.
  }\label{fig:cls_gs_diff_embed_comparison}

@@

% TODO?: RandomForest performence on sparse embeddings

@@

  \caption{Comparison of F1 score per each transformation of best classifiers
  for \Embed{sbert} and \Embed{tfidf} embedding in contextualized
  setting.}\label{fig:cls_gs_diff_labels}
\end{figure}

@@

\subsubsection{No-context setting}

Grid search results for no-context setting are displayed in
Figure~\ref{fig:cls_gs_no_context_embed_comparison}. Comparing these results to
those in contextualized setting, we see large drop of performance for
\Embed{sbert} embedding --- almost by 50\%. This shows how much the classifiers
relied on the information contained in the embedding of seed sentence. This
drop is not as pronounced for \Embed{sbert\_supervised} embedding, since the
model producing it is optimized for good scores in precisely this type of
task.\footnote{For this reason, we avoid comparing \Embed{sbert\_supervised} to
other embeddings from this point onwards.} On the other hand, the BOW
embeddings performed almost as well, suggesting they were not really able to
take advantage of the contextualized setting.

@@

  \caption{Grid search F1 macro averaged scores of sentence embeddings with no
  context. The reported scores of \Embed{sbert\_supervised} are the mean scores
  of all SBERT supervised
  embeddings.}\label{fig:cls_gs_no_context_embed_comparison}

@@

In Figure~\ref{fig:cls_gs_no_context_labels} we compare performance of best
classifiers for \Embed{sbert} and \Embed{tfidf} embedding. Whereas the most
recognised transformations are the same as in the contextualized setting, the
classifiers performance significantly drops in all cases except for the
\Trans{ban} and the \Trans{possibility} transformations embedded with TF-IDF.
This suggests that \Trans{ban} and \Trans{possibility} transformations can be
easily identified from pure \emph{lexical} features. While SBERT fails to
encode this information into its embedding, it performs even better when the
embeddings of a sentence and its corresponding seed sentence are put into one
context. This phenomena is pronounced mainly for \Trans{different meaning} and
\Trans{past} transformations, where the inclusion of seed sentence's embedding
on the input is responsible for over 80\% and 60\% of F1 score of the
classifier, as can be seen in Figure~\ref{fig:cls_gs_labels_context_diff}.
Overall SBERT's embeddings benefit from contextualization more than TF-IDF
vectors.

@@

  \caption{Comparison of F1 score per each transformation of best classifiers
  for \Embed{sbert} and \Embed{tfidf} embedding in no-context
  setting.}\label{fig:cls_gs_no_context_labels}

@@

  \caption{Difference in F1 scores between no-context and contextualized
  setting as a percentage of contextualized
  score.}\label{fig:cls_gs_labels_context_diff}.

>>>>

@@

% David: I would maybe skip the part about training shorter embeddings. But
% that can wait for the paper submit.
For training the Doc2vec model, we have used the implementation of Doc2vec
algorithm available in the Python package gensim, using a window of size 4 and
considering every word that occured at all (no pruning of rare words) and the
vector size of 512. We have also tried training Doc2Vec models with different
embedding sizes (2, 3, 10, 100, 256), partly so that we could see, ehat impact
the size change has on the semantic label predictability, partly (especially if
we detected a positive correlation between the embedding size and the semantic
label predictability) in order to try and get hint as to how prominent the
specific semantic features are (i.e. which is the smallest vector size at which
given semantic label is predictable from the corresponding embeddings).

@@

The motivation for these tasks is that if the embedding contains semantic
information (in our case the type of transformation used), it would be
beneficial for the classifier to find it and use it. This would boost its
scores and make it standout.

Since transformations define the relationship of two embeddings, we hypothesize
that classifiers need both the sentence embedding and the seed embedding to
produce reliable predictions. To test our hypothesis we run the probing tasks
twice: first classifying transformation vectors\footnote{We also tried
concatenating the sentence and the seed embeddings. This approach produced
overall worse results, and therefore we omit them.}, second classifying only
the sentence embeddings themselves. The results are presented in
Section~\ref{sec:probing_results}, the complete method is described in detail
in Appendix~\ref{appendix:probing_method}.

@@

\subsubsection{Classifying transformation vectors}

@@

  \caption{Grid search F1 macro averaged scores for transformation vectors.
  The reported scores of \Embed{sbert\_supervised} are the mean scores of all
  SBERT supervised embeddings.
  }\label{fig:cls_gs_diff_embed_comparison}

@@

% TODO?: Should I mention the suprising performance of RandomForest on sparse
% embeddings?

@@

  \caption{F1 score of transformation vectors for each transformation of the
  best classifiers for given embedding.}\label{fig:cls_gs_diff_labels}

@@

\subsubsection{Classifying sentence embeddings}

Again, we report plotted grid search results in
Figure~\ref{fig:cls_gs_no_context_embed_comparison}. Comparing these results to
those in contextualized setting, we see large drop of performance for
\Embed{sbert} embedding --- almost by 50\%. This shows how much the classifiers
relied on the information contained in the embedding of seed sentence. This
drop is not as pronounced for \Embed{sbert\_supervised} embedding, because the
model producing them is optimized with precisely this type of
task.\footnote{For this reason, we avoid comparing \Embed{sbert\_supervised} to
other embeddings from this point onwards.} On the other hand, the BOW
embeddings performed almost as well, suggesting they were not really able to
take advantage of the contextualized setting.

@@

  \caption{Grid search F1 macro averaged scores of sentence embeddings. The
  reported scores of \Embed{sbert\_supervised} are the mean scores
  of all SBERT supervised
  embeddings.}\label{fig:cls_gs_no_context_embed_comparison}

@@

In Figure~\ref{fig:cls_gs_no_context_labels} we compare performance of best
classifiers for \Embed{sbert} and \Embed{tfidf} embedding. Whereas the most
recognised transformations are the same as when classifying transformation
vectors, the classifiers performance significantly drops in all cases except
for the \Trans{ban} and the \Trans{possibility} transformations embedded with
TF-IDF. This suggests that \Trans{ban} and \Trans{possibility} transformations
can be easily identified from pure \emph{lexical} features and based on the
sentence embedding only. While SBERT fails to encode this information into its
embedding, it performs even better than TF-IDF when the embeddings of a
sentence and its corresponding seed sentence are put into one context. As can
be seen in Figure~\ref{fig:cls_gs_labels_context_diff}, this phenomena is
pronounced mainly for \Trans{different meaning} and \Trans{past}
transformations, where the inclusion of seed sentence's embedding on the input
is responsible for over 80\% and 60\% of F1 score of the classifier. Overall
SBERT's embeddings benefit from including the seed embedding to the input more
than TF-IDF vectors.

@@

  \caption{F1 scores of sentence embeddings for each transformation of the best
  classifiers for given embedding.}\label{fig:cls_gs_no_context_labels}

@@

  \caption{Drop in performance of best classifiers for given embedding going
  from transformation vector to sentence embedding classification. The drop is
  given as a percentage of score for transformation
  vectors.}\label{fig:cls_gs_labels_context_diff}.
